Physics-based protein structure prediction and design using the confinement method

### 2013-01-09 Paper/paper.tex
Casp

This method is particularly well suited to pick out the native/native like structure from misfolded or decoys. We have
tested this using different models from CASP (critical assessment of structure
prediction). CASP is a blind test in which different groups world wide apply their
methods to predict the 3D structure of proteins given their sequence. This is done with a strict 3
week limit on each target (3 day for servers) and each group is allowed to submit 5 possible
structures (ranked from better to worst). In our role as assessors during the CASP8 and CASP9 \cite{MacCallum2011}
competition we observed no correlation at all between the ordering of structures submited by the
groups and the real ranking compared to an experimental model \cite{Kryshtafovych2011}. The consequences of this go beyond
those five structures; the deeper meaning is that groups producing ensembles of structures are
generating structures that are better than the ones they submit, but they do not know about it. In
fact, when querrying different groups after the results were known, most groups agree that this is
the case. Beyond the CASP problem, this reflects on the modelers ability to correctly rank order
models in many different environments, from structures for drug design leads to designing more
stable proteins or peptide mimetics such as peptoids. One of the main culprits of this lack of
accuracy is the fact that ranking is often done via a potential energy function, which in many cases
lacks an entropy component. Other initiatives including knowledge based potentials do use some sort
of free energy to rank order, but its accuracy is not enough. Our method provides a physics based
solution to this problem. We have tried two experiments centered on CASP. First we tried to rank
order some structures from the previous CASP9 experiment. We have tested different options for that purpose.
First, we rank order submition from same group. Then we rank submition from different groups. Our second test was 
to check whether this method can do better quality assessment/metaprediction than the best group in that catagory in CASP9. 
In both methods, our ranking is determined as a free energy measure,
and it is compared with the ranking given by a geometrical comparison between native and submitted
models called GDT\_TS (Global distance test score \cite{Zemla2003}). 
GDT\_TS represents the average percentage of residues that are in close proximity in two
structures optimally superimposed using four different distance cutoffs (1, 2, 4 and 8 Å).

We have chosen the inital models on which to test the methods based on server groups that have
traditionally done good in past CASP events. There is a high correlation between the structures that
are chosen for this method and its predictive accuracy. In particular, when GDT\_TS scores are below
50, it is difficult for us to say anything about the models. Our approach is simple, with every
subset of structures we choose, we use the confinement method in a pairwise procedure, and then rank
the structures. In all cases, we have calculated the free energy with respect to the native structure either obtained from NMR or 
x-ray crystallography but we do not need to make comparisons to native in order to do rank ordering.

\subsubsection{confinement method can rank different models submitted by the same group and distinguish the native structure}

The question that we want to answer here is whether we can rank models submitted by the same group.In all cases, we have calculated the free energy
difference between the native structure and the submitted models as well. 
%Identifying the native state from a set of models is a very easy test, that most metapredictors
%already handle correctly, but it is also a first proof of principle. 
First we choose a protein BVU3908 from Bacteroides vulgatus whose PDB id and CASP target code are
2L01 and T0559 respectively. The native structure of this 69 residue protein was solved using NMR.
The best predictor group for this target was "BAKER-ROSETTASERVER".
We initlally checked similarity between the models submitted by "BAKER-ROSETTASERVER" and discarded two of them from the 
analysis on the basis of being too similar to some of the other models. The GDT\_TS score and rmsd
values as shown in Figure~\ref{fig:T0559} indicate that the model 1 was predicted correctly, 
whereas the order of model 3 and 5 was wrong. The main difference
between the model 3 and the rest of the model is that the orientation of first alpha helix of model 3 is opposite.
On the other hand, as shown in Figure~\ref{fig:T0559} confine and release method not only can differentiate the native 
structure from the submitted models , it can also correctly rank all the models like their The GDT\_TS score.

To further test the method we have taken the example of protein BT2368 from Bacteroides thetaiotaomicron. The PDB id of this 
74 amino acid residue protein is 2L02 and CASP target code is T0560. We have compared the free energy difference between 
the native structure and the two of the five submitted models from the group "Splicer". The remaining three models are discarded as 
again they are similar to the rest of the models. The comparison of GDT\_TS score as shown in 
Figure~\ref{fig:T0560} and published in final CASP result is only for models with residues 3 to 66. In order to keep
consistency, we have also done our analysis with models and crystal structure consisting residues 3 to 66. As expected,
the native state is identified correctly, and the two other models are ranked in the correct order in accordance with the GDT\_TS score.

\subsubsection{Confinement method can rank models submitted by different groups}

Our next test was between models from the best submissions of different groups. 
For this purpose we have chosen the x-ray crystallographic structure of 
fas apoptosis inhibitory protein molecule whose pdb id and CASP target codes are 3MX7 and T0540 respectively.
This protein contain 90 amino acid residues with 8 beta strands. We have chosen best models from groups
"LTB" and "Mufold" for our analysis, which are labelled as Model 1 and Model 2 in Figure~\ref{fig:T0540}.
As presented in  Figure~\ref{fig:T0540} we found that this method is able to correctly order the models which matches
with the GDT\_TS score.

\subsubsection{Per residue free energy calculation identify the residues that are responsible for difference between two conformation}

In order to understand the per residue contribution towards the free energy, we have first calculated the total free energy of NMR 
structure of a domain of adhesion exoprotein from Pediococcus pentosaceus and its submitted CASP9 model. The 
pdb id and CASP code of this 79 residue protein are 2KWY and T0569 respectively. We have compared the native structure with the
best model structure submitted for this target in CASP9. This model structure was submitted by the group " Mufold". All the
five model structures submitted by this group was found to be close, within a GDT\_TS score  of 72 to 78. Thus we proceed by 
comparing only the native structure and
the best predicted model of this target. Visualization upon superimposition of two structures reveals that
the difference between the two structure is at the region consist of residues 48 to 65, where model1 is much more disordered compare to the
native structure(see Figure~\ref{fig:T0569}). Other regions look similar for both the cases with of model1 is within 2.6 \AA backbone
rmsd value of the native structure. We found that, confinement method can still differentiate between the native structure and 
the generated model. We have also calculated the per residue free energy with a aim that this can identify the residue which are responsible
for such free enerfy difference. The per residue free energy decomposition is shown in  Figure~\ref{fig:T0569_per_residue}(A). We indeed 
found that two hydrophobic residues Val-59 and Ile-61 are oriented towards the solvent in the protein 
surface in the generated model (Figure~\ref{fig:T0569_per_residue}(B). This the region, where there is a beta sheet in the original crystal structure, which is absent in the 
model. There are also some region, which stabilizes the generated model compare to that of the crystal. A close inspection reveal that 
the hydrophilic Glu in position 67 is completely exposed to the protein surface and solvent in the best model, whereas it is partially exposed in the 
crystal. This is presented in Figure~\ref{fig:T0569_per_residue}(C). The peak at position 76 is due to the H-bond between Lys-76 and Asp-11, 
which is absent in original crystalographic structure (Figure~\ref{fig:T0569_per_residue}(D)).

>>>>

This method is particularly well suited to pick out the native/native like structure from misfolded or decoys. We have
tested this using different models from CASP (critical assessment of structure
prediction). CASP is a blind test in which different groups world wide apply their
methods to predict the 3D structure of proteins given their sequence. This is done with a strict 3
week limit on each target (3 day for servers) and each group is allowed to submit 5 possible
structures (ranked from best to worst). In our role as assessors during the CASP8 and CASP9 \cite{MacCallum2011}
competition we observed no correlation at all between the ordering of structures submited by the
groups and the real ranking compared to an experimental model \cite{Kryshtafovych2011}. The
consequences of this goes beyond
those five structures; the deeper meaning is that groups producing ensembles of structures are
generating structures that are better than the ones they submit, but they do not know about it. In
fact, when querrying different groups after the results were known, most groups agree that this is
the case. Beyond the CASP problem, this reflects on the modelers ability to correctly rank order
models in many different environments, such as structures for drug design leads or designing more
stable proteins or peptide mimetics such as peptoids. One of the main culprits of this lack of
accuracy is the fact that ranking is often done via a potential energy function, which in many cases
lacks an entropy component. Other initiatives including knowledge based potentials do use some sort
of free energy to rank order, but its accuracy is not enough. Our method provides a physics based
solution to this problem. We have tried two experiments centered on CASP. First we tried to rank
order some structures from the previous CASP9 experiment. We have tested different options for that purpose.
First, we rank order submitions from same group and then from different groups. Our second test was 
to check whether this method can do better quality assessment/metaprediction than the best group in that catagory in CASP9. 
In both methods, our ranking is determined as a free energy measure,
and it is compared with the ranking given by a geometrical comparison between native and submitted
models called GDT\_TS (Global distance test score \cite{Zemla2003}). 
GDT\_TS represents the average percentage of residues that are in close proximity in two
structures optimally superimposed using four different distance cutoffs (1, 2, 4 and 8 Å).

We have chosen the inital models on which to test the methods based on server groups that have
traditionally done good in past CASP events. There is a high correlation between the structures that
are chosen for this method and its predictive accuracy. In particular, when GDT\_TS scores are below
50, it is difficult for us to say anything about the models. Our approach is simple, for every
subset of structures we choose, we use the confinement method in a pairwise procedure, and then rank
the structures. In all cases, we have also calculated the free energy with respect to the native
structure to check the stability of the models.
However, this step is not needed if we just want to rank order.

\subsubsection{The confinement method can rank different models generated by the same methodology and distinguish the native structure}

The question that we want to answer here is whether we can rank models that have been generated with
the same methodology. In all cases, we have calculated the free energy
difference between the native structure and the submitted models as well. 
%Identifying the native state from a set of models is a very easy test, that most metapredictors
%already handle correctly, but it is also a first proof of principle. 
First we choose a protein BVU3908 from Bacteroides vulgatus whose PDB id and CASP target code are
2L01 and T0559 respectively. The native structure of this 69 residue protein was solved using NMR.
The best predictor group for this target was "BAKER-ROSETTASERVER".
We initially checked similarity between the models submitted by "BAKER-ROSETTASERVER" and discarded two of them from the 
analysis on the basis of being too similar to some of the other models. The GDT\_TS score and rmsd
values as shown in Figure~\ref{fig:T0559} indicate that model 1 was predicted correctly, 
whereas the order of model 3 and 5 was wrong. The main difference
between model 3 and the rest of the models is that the orientation of the first alpha helix of model
3 is \Alberto{opposite??}.
On the other hand, as shown in Figure~\ref{fig:T0559} the confine and release method not only can differentiate the native 
structure from the submitted models, the ranking also correlates perfectly with the GDT\_TS score.

To further test the method we have taken the example of protein BT2368 from Bacteroides thetaiotaomicron. The PDB id of this 
74 amino acid residue protein is 2L02 and CASP target code is T0560. We have compared the free energy difference between 
the native structure and the two of the five submitted models from the group "Splicer". The remaining three models are discarded as 
again they are similar to the rest of the models. The comparison of GDT\_TS score as shown in 
Figure~\ref{fig:T0560} and published in the final CASP result is only for models with residues 3 to 66. In order to keep
consistency, we have also done our analysis with models and crystal structure consisting residues 3 to 66. As expected,
the native state is identified correctly, and the two other models are ranked in the correct order in accordance with the GDT\_TS score.

\subsubsection{The confinement method can rank models generated by different methods}

Our next test was between models that are produced with different methodologies. We do this by
looking at some of the best CASP submissions from different research groups. 
For this purpose we have chosen the x-ray crystallographic structure of 
fas apoptosis inhibitory protein molecule whose pdb id and CASP target codes are 3MX7 and T0540 respectively.
This protein contain 90 amino acid residues with 8 beta strands. We have chosen best models from groups
"LTB" and "Mufold" for our analysis, which are labelled as Model 1 and Model 2 in Figure~\ref{fig:T0540}.
As presented in  Figure~\ref{fig:T0540} we found that this method is able to correctly order the models which matches
with the GDT\_TS score.

\subsubsection{Per residue free energy calculation identifies the residues that are responsible for
differences between two conformations}

We used one of the CASP targets (pdb id=2KWY, casp id=T0569) to study the per residue contribution
between this NMR structure and the best CASP9 predicted model (from the 'Mufold' group, GDT\_TS=78).
Visualization upon superimposition of two structures reveals that
the difference between the two structure is at the region consist of residues 48 to 65, where model1 is much more disordered compare to the
native structure(see Figure~\ref{fig:T0569}). Other regions look similar for both the cases with of model1 is within 2.6 \AA backbone
rmsd value of the native structure. We found that, the confinement method can still differentiate between the native structure and 
the generated model. We have also calculated the per residue free energy with a aim that this can identify the residue which are responsible
for such free enerfy difference. The per residue free energy decomposition is shown in  Figure~\ref{fig:T0569_per_residue}(A). We indeed 
found that two hydrophobic residues Val-59 and Ile-61 are oriented towards the solvent in the protein 
surface in the generated model (Figure~\ref{fig:T0569_per_residue}(B). This the region, where there is a beta sheet in the original crystal structure, which is absent in the 
model. Interestinlgy, some regions are more stable in the generated model. A close inspection
reveals that 
the hydrophilic Glu in position 67 is completely exposed to the protein surface and solvent in the best model, whereas it is partially exposed in the 
crystal (Figure~\ref{fig:T0569_per_residue}(C)). The peak at position 76 is due to the H-bond between Lys-76 and Asp-11, 
which is absent in the original crystalographic structure (Figure~\ref{fig:T0569_per_residue}(D)).
